---
title: "Choropeth Mapping with R"
subtitle: "Hands-on Exercise 2"
execute:
  warning: False  # This turns off warning messages being displayed
date: 8 January, 2024
date-modified: "last-modified"
---

# Overview

Thematic mapping involves the use of map symbols to visualize selected properties of geographic features that are not naturally visible, such as population, temperature, crime rate, and property prices, just to mention a few of them.

GeoVisualization, on the other hand, works by providing graphical ideation to render a place, a phenomenon or a process visible, enabling human's most powerful information-processing abilities - those of spatial cognition associated with our eye-brain vision system - to be directly brought to bear.

In this chapter, you will learn how to plot functional and truthful choropleth maps by using R package called **tmap** package.

# Getting Started

In this exercise, the key R package used is **tmap** package in R. Besides **tmap** package, four other R packages will be used. They are:

-   **readr** for importing delimited text file

-   **tidyr** for tidying data

-   **dplyr** for wrangling data

-   **sf** for handling geospatial data

Among the four packages, **readr**, **tidyr** and **dplyr** are part of **tidyverse** package.

The code below will be used to install and load these packages in RStudio.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

Notice we only need to install **tidyverse** instead of **readr**, **tidyr** and **dplyr** individually!

# Importing Data into R

## The Data

Two data set will be used to create the choropeth map.

-   Master Plan 2014 Subzone Boundary (Web) (i.e. `MP14_SUBZONE_WEB_PL`)

    -   ESRI shapefile format (Geospatial data)

    -   Downloaded from [data.gov.sg](https://data.gov.sg)

    -   Consists of the geographical boundary of Singapore at the planning subzone level.

    -   The data is based on URA Master Plan 2014

-   Singapore Residents by Planning Area / Subzone, Age Group, Sex and Type of Dwelling, June 2011-2020 (i.e. `respopagesextod2011to2020.csv`)

    -   CSV format (Aspatial data)

    -   Downloaded from [Department of Statistics, Singapore](https://singstat.gov.sg)

    -   Although it does not contain any coordinate values, it's PA and SZ fields can be used as unique identifiers to geocode to `MP14_SUBZONE_WEB_PL` shapefile.

## Importing Geospatial Data

The code below uses the `st_read()` function of **sf** package to import `MP14_SUBZONE_WEB_PL` shapefile into R as a simple feature data frame called `mpsz`.

```{r}
mpsz <- st_read(dsn = "../../data/geospatial/MP14_SUBZONE_WEB_PL",
                layer = "MP14_SUBZONE_WEB_PL")
mpsz
```

You can examine the content of `mpsz` by using the code below.

```{r}
mpsz
```

Notice that only the first ten records will be displayed!

## Importing Attribute Data

We will import `repopagesextod2011to2020.csv` into RStudio and save the file into an R dataframe called *popdata.*

The task will be performed by using `read_csv()` from **readr** package as show below.

```{r}
popdata <- read_csv("../../data/aspatial/respopagesextod2011to2020/respopagesextod2011to2020.csv")
```

## Data Preparation

Before a thematic map can be prepared, we must prepare a data table with year 2020 values. The data table should include the variables PA, SZ, YOUNG, ECONOMY ACTIVE, AGED, TOTAL, DEPENDENCY.

-   **YOUNG**: age group 0 to 4 until age group 20 to 24

-   **ECONOMY ACTIVE**: age group 25-29 until age group 60-64

-   **AGED**: age group 65 and above

-   **TOTAL**: all age group

-   **DEPENDENCY**: the ratio between young and aged against economy active group

#### Data Wrangling

The following data wrangling and transformation functions will be used:

-   `pivot_wider()` of **tidyr** package

-   `mutate()`, `filter()`, `group_by()` and `select()` of **dplyr** package

```{r}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup()%>%
  pivot_wider(names_from=AG, 
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>%
mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+
rowSums(.[13:15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%
mutate(`TOTAL`=rowSums(.[3:21])) %>%  
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
/`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, 
       `ECONOMY ACTIVE`, `AGED`, 
       `TOTAL`, `DEPENDENCY`)
```

#### Joining the Attribute data and Geospatial data

Before we can perform the georelational join, one extra step is required to convert the values in PA and SZ fields to uppercase. This is because the values of PA and SZ fields are made up of upper- and lowercase. On the other, hand the SUBZONE_N and PLN_AREA_N are in uppercase.

```{r}
popdata2020 <- popdata2020 %>%
  mutate_at(.vars = vars(PA, SZ), 
          .funs = list(toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

Next, `left_join()` of **dplyr** is used to join the geographical data and attribute table using planning subzone name e.g. *SUBZONE_N* and *SZ* as the common identifier.

```{r}
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))
```

Thing to learn from the code chunk above:

-   `left_join()` of **dplyr** package is used with `mpsz` simple feature data frame as the left data table is to ensure that the output will be a simple features data frame.

```{r}
write_rds(mpsz_pop2020, "../../data/rds/mpszpop2020.rds")
```

# Choropleth Mapping Geospatial Data using `tmap`

Choropleth mapping involves the symbolization of enumeration units, such as countries, provinces, states, counties or census units, using area patterns or graduated colors. For example, a social scientist may need to use a choropleth map to portray the spatial distribution of aged population of Singapore by Master Plan 2014 Subzone Boundary.

Two approaches can be used to prepare thematic map using **tmap**, they are:

-   Plotting a thematic map quickly using `qtm()`.

-   Plotting highly customization thematic map by using **tmap** elements.

## Plotting a Choropleth map quickly using `qtm()`

The easiest and quickest to draw a choropleth map using **tmap** is using `qtm()`. It is concise and provides a good default visualization in many cases.

The code below will draw a cartographic standard choropleth map as shown below.

```{r}
tmap_mode("plot")
qtm(mpsz_pop2020,
    fill = "DEPENDENCY")
```

Things to learn from the code chunk above:

-   `tmap_mode()` with "plot" option is used to produce a static map. For interactive mode, "view" option should be used.

-   `fill` argument is used to map the attribute (i.e. DEPENDENCY)

## Creating a Choropleth map by using `tmap`'s elements

Despite its usefulness of drawing a choropleth map quickly and easily, the disadvantage of `qtm()` is that it makes aesthetics of individual layers harder to control. To draw a high quality cartographic choropleth map, **tmap**'s drawing elements should be used.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          title = "Dependency ratio") +
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45,
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) + 
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(alpha=0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authority (URA)\n and Population data from Department of Statistics DOS",
             position = c("left", "bottom"))
```

In the following sections, we will break down the process for plotting these elements.

### Drawing a base map

The basic building block of **tmap** is `tm_shape()` followed by one or more layer elements such as `tm_fill()` and `tm_polygons()`.

In the code below,

-   `tm_shape()` is used to define the input data (i.e. `mpsz_pop2020`) and

-   `tm_polygons()` is used to draw the planning subzone polygons.

```{R}
tm_shape(mpsz_pop2020) +
  tm_polygons()
```

### Drawing a choropleth map using `tm_polygons()`

To draw a choropleth map showing the geographical distribution of a selected variable by planning subzone, we just need to assign the target variable such as *Dependency* to `tm_polygons()`.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY")
```

Things to learn from `tm_polygons()`:

-   The default interval binning used to draw the choropleth map is called "pretty".

-   The default color scheme used is `YlOrRd` of ColorBrewer.

-   By default, missing values will be shaded in grey.

### Drawing a choropleth map using `tm_fill()` and `tm_border()`

Actually, `tm_polygons()` is a wrapper of `tm_fill()` and `tm_border()`.

-   `tm_fill()` shades the polygons by using the default color scheme

-   `tm_borders()` adds the borders of the shapefile onto the choropleth map

The code below draws a choropleth map by using the `tm_fill()` alone.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY")
```

Notice that the planning subzones are shared according to the respective dependency values.

To add the boundary of the planning subzones, `tm_borders()` will be used in the code below.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY") +
  tm_borders(lwd = 0.1, alpha = 1)
```

Notice that light-gray border lines have been added.

The `alpha` argument is used to define transparency number between 0 (transparent) and 1 (opaque). By default, the alpha value of the col is used (normally 1).

Besides `alpha` argument, there are three other arguments for `tm_borders()`, they are:

-   `col` = border color.

-   `lwd` = border line width. The default is 1.

-   `lty` = border line type. The default is "solid".

## Data Classification methods of `tmap`

Most choropleth maps employ some methods of data classification. The point of classification is to take a large number of observations and group them into data ranges or classes.

**tmap** provides a total of 10 data classification methods: `fixed`, `sd`, `equal`, `pretty` (default), `quantile`, `kmeans`, `hclust`, `bclust`, `fisher`, and `jenks`.

To define a data classification method, the `style` argument of `tm_fill()` or `tm_polygons()` will be used.

### Plotting choropleth maps with built-in classification methods

The code below shows a `quantile` data classification that uses 5 classes.

```{r}
tm_shape(mpsz_pop2020) + 
  tm_fill("DEPENDENCY",
          n = 5,
          style = "jenks") +
  tm_borders(alpha = 0.5)
```

In the code below, `equal` data classification method is used.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          n = 5,
          style = "equal") + 
  tm_borders(alpha = 0.5)
```

Notice that the distribution of `quantile` data classification method are more evenly distributed than `equal` data classification method.

::: callout-warning
## Warning: Maps Lie!

DIY: Using what you had learned prepare choropleth maps by using different classification methods supported by tmap and compare their differences

DIY: Preparing choropleth maps by using similar classification method but with different numbers of classes (i.e. 2, 6, 10, 20). Compare the output maps, what observation can you draw?

See the bottom of this page for the maps.
:::

### Plotting choropleth map with custom break

For all the built-in styles, the category breaks are computed internally. In order to override these defaults, the breakpoints can be set explicitly by means of the `breaks` argument in `tm_fill()`. It is important to note that, in **tmap**, the breaks include a minimum and maximum. As a result, in order to end up with n categories, n+1 elements must be specified in the `breaks` option (the values must be in increasing order).

Before we get started, it is always a good practice to get some descriptive statistics on the variable before setting the break points. The code below will be used to compute and display the descriptive statistics of **DEPENDENCY** field.

```{r}
summary(mpsz_pop2020$DEPENDENCY)
```

With reference to the results above, we set break point at 0.60, 0.70, 0.80, and 0.90. In addition, we also need to include a minimum and maximum, which we set at 0 and 100. Our `breaks` vector is thus c(0, 0.60, 0.70, 0.80, 0.90, 1.00)

Now, we will plot the choropleth map using the code below.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  tm_borders(alpha = 0.5)
```

## Color Scheme

**tmap** supports color ramps either defined by the user or a set of predefined color ramps from the **RColorBrewer** package.

### Using ColorBrewer palette

To change the color, we assign the preferred color to palette argument of `tm_fill()` as shown below.

```{r}
tm_shape(mpsz_pop2020) + 
  tm_fill("DEPENDENCY",
          n = 6,
          style = "quantile",
          palette = "Blues") +
  tm_borders(alpha = 0.5)
```

To reverse the color shading, add a "-" prefix.

```{r}
tm_shape(mpsz_pop2020) + 
  tm_fill("DEPENDENCY",
          n = 6,
          style = "quantile",
          palette = "-Blues") +
  tm_borders(alpha = 0.5)
```

## Map Layouts

Map Layouts refers to the combination of all map elements into a cohesive map. Map elements include among others the objects to be mapped, the title, the scale bar, the compass, margins and aspect ratios. Colour settings and data classification methods covered in the previous section relate to the palette and break-points are used to affect how the map looks.

### Map Legend

In **tmap**, several `legend` options are provided to change the placement, format and appearance of the legend.

```{r}
tm_shape(mpsz_pop2020) + 
  tm_fill("DEPENDENCY",
          n = 6,
          style = "jenks",
          palette = "Blues",
          legend.hist = TRUE,
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone \n(Jenks Classification)",
            main.title.position = "center",
            main.title.size = 1,
            legend.height = 0.45,
            legend.width = 0.35,
            legend.outside = FALSE,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```

### Map Style

**tmap** allows a wide variety of layout settings to be changed. They can be called by using `tmap_style()`

The code chunk below shows the `classic` style is used.

```{r}
tm_shape(mpsz_pop2020) + 
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "-Greens",) +
  tm_borders(alpha = 0.5) +
  tmap_style("classic")
```

### Cartographic Furniture

Besides map style, **tmap** also provides arguments to draw other map furniture such as compass, scale bar and grid lines.

In the code chunk below, `tm_compass()`, `tm_scale_bar()` and `tm_grid()` are used to add compass, scale bar and grid lines onto the choropleth map.

```{r}
tm_shape(mpsz_pop2020) + 
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          title = "No. of persons") +
  tm_layout(main.title = "Distribution of Dependency Ratio \nby planning subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45,
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scale_bar(width = 0.15) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authority (URA)\n and Population data from Department of Statistics DOS",
             position = c("left", "bottom"))
```

To reset the default style, refer to the code below.

```{r}
tmap_style("white")
```

## Drawing Small Multiple Choropleth Maps

**Small multiple maps**, also referred to as **facet maps**, are composed of many maps arrange side-by-side, and sometimes stacked vertically. Small multiple maps enable the visualization of how spatial relationships change with respect to another variable, such as time.

In **tmap**, small multiple maps can be plotted in 3 ways:

-   by assigning multiple values to at least one of the aesthetic arguments

-   by defining a group-by variable in `tm_facets()`

-   by creating multiple stand-alone maps with `tmap_arrange()`

### By assigning multiple values to at least one of the aesthetic arguments

In this example, small multiple choropleth maps are created by defining `ncols` in `tm_fill()`

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill(c("YOUNG", "AGED"),
          style = "equal",
          palette = "Blues") +
  tm_layout(legend.position = c("right", "bottom")) +
  tm_borders(alpha = 0.5) +
  tmap_style("white")
```

In this example, small multiple choropleth maps are created by assigning multiple values to at least one of the aesthetic arguments

```{r}
tm_shape(mpsz_pop2020) + 
  tm_polygons(c("DEPENDENCY","AGED"),
              style = c("equal", "quantile"),
              palette = list("Blues","Greens")) +
  tm_layout(legend.position = c("right", "bottom"))
```

### By defining a group-by variable in `tm_facets()`

In this example, multiple small choropleth maps are created by using `tm_facets()`

```{r}
tm_shape(mpsz_pop2020) + 
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          thres.poly = 0) +
  tm_facets(by = "REGION_N",
            free.coords = TRUE,
            drop.shapes = TRUE) +
  tm_layout(legend.show = FALSE,
            title.position = c("center", "center"),
            title.size = 20) +
  tm_borders(alpha = 0.5)
```

### By creating multiple stand-alone maps with `tmap_arrange()`

In this example, multiple small choropleth maps are created by creating multiple stand-alone maps with `tmap_arrange()`

```{r}
youngmap <- tm_shape(mpsz_pop2020) +
  tm_polygons("YOUNG",
              style = "quantile",
              palette = "Blues")

agedmap <- tm_shape(mpsz_pop2020) +
  tm_polygons("AGED",
              style = "quantile",
              palette = "Blues")

tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```

## Mapping Spatial Object Meeting a Selection Criterion

Instead of creating small multiple choropleth map, you can also use selection function to map spatial objects meeting the selection criterion.

```{r}
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N=="CENTRAL REGION", ]) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          legend.hist = TRUE,
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(legend.outside = TRUE,
            legend.height = 0.45,
            legend.width = 5.0,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```

# DIY Exercise

DIY: Using what you had learned prepare choropleth maps by using different classification methods supported by tmap and compare their differences

DIY: Preparing choropleth maps by using similar classification method but with different numbers of classes (i.e. 2, 6, 10, 20). Compare the output maps, what observation can you draw?

```{r}
quantile_lowbin_map <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              n = 5,
              style = "quantile",
              palette = "Blues")

quantile_highbin_map <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              n = 15,
              style = "quantile",
              palette = "Blues")

equal_lowbin_map <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              n = 5,
              style = "equal",
              palette = "Greens")

equal_highbin_map <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              n = 15,
              style = "equal",
              palette = "Greens")

tmap_arrange(quantile_lowbin_map, 
             equal_lowbin_map, 
             quantile_highbin_map, 
             equal_highbin_map, 
             ncol=2)
```

Observations:

-   Different classification methods and bin numbers can make the distribution look more polarizing, which allows the narrative to emphasize the difference between the lowest and highest parts of the distribution.

# Reference

## All about `tmap` package

-   [tmap: Thematic Maps in R](https://jstatsoft.org/article/view/v084i06)

-   [tmap](https://cran.r-project.org/web/packages/tmap/index.html)

-   [tmap: get started!](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html)

## Geospatial Data Wrangling

-   [sf: Simple Features for R](https://cran.r-project.org/web/packages/sf/index.html)

-   [Simple Features for R: Standardized Support for Spatial Vector Data](https://journal.r-project.org/archive/2018/RJ-2018-009/RJ-2018-009.pdf)

-   [Reading, Writing and Converting Simple Features](https://cran.r-project.org/web/packages/sf/vignettes/sf2.html)

## Data Wrangling

-   [dplyr](https://dplyr.tidyverse.org/)

-   [Tidy data](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html)

-   [tidyr: Easily Tidy Data with `'spread()'` and `'gather()'` Functions](https://cran.r-project.org/web/packages/tidyr/tidyr.pdf)
