---
title: "R for Geospatial Data Science"
subtitle: "In-class Exercise 2"
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    toc: true # Table of Contents
execute:
  eval: true
  echo: true
  warning: False  # This turns off warning messages being displayed
date: 15 January, 2024
date-modified: "last-modified"
---

# Getting Started

The following R packages will be used:

-   arrow

    -   To handle cross-language development platform to handle databases and data conversion. (e.g. read Parquet files)

-   lubridate

    -   To work with date and time. Part of the tidyverse family, but needs to be installed independently.

-   tidyverse

    -   A family of R packages commonly used for data science work

-   tmap

-   sf

```{r}
pacman::p_load(arrow, lubridate, tidyverse, tmap, sf)
```

# Importing Grab-Posisi Dataset

```{r}
df <- read_parquet("../../data/geospatial/GrabPosisi/part-00000-8bbff892-97d2-4011-9961-703e38972569.c000.snappy.parquet")
df
```

# Converting into sf tibble data.frame

```{r}
#sfdf <- st_as_sf(df)
#sfdf
```

# Visualizing as Point Symbol Map

# Importing Master Plan 2019 Planning Subzone Boundary
