---
title: "R for Geospatial Data Science"
subtitle: "In-class Exercise 2"
format:
  html:
    code-fold: false
    code-summary: "Show the code"
    toc: true # Table of Contents
execute:
  eval: true
  echo: true
  warning: False  # This turns off warning messages being displayed
date: 15 January, 2024
date-modified: "last-modified"
---

# Getting Started

The following R packages will be used:

-   arrow

    -   To handle cross-language development platform to handle databases and data conversion. (e.g. read Parquet files)

-   lubridate

    -   To work with date and time. Part of the tidyverse family, but needs to be installed independently.

-   tidyverse

    -   A family of R packages commonly used for data science work

-   tmap

-   sf

```{r}
pacman::p_load(arrow, lubridate, tidyverse, tmap, sf)
```

# Importing Grab-Posisi Dataset

Grab-Posisi dataset is in parquet file format, we will use the **arrow** package to import it.

```{r}
#| eval: false
df <- read_parquet("../../data/geospatial/GrabPosisi/part-00000-8bbff892-97d2-4011-9961-703e38972569.c000.snappy.parquet")
```

Let's convert the pingtimestamp column from integer to a datetime format.

```{r}
#| eval: false
df$pingtimestamp <- as_datetime(df$pingtimestamp)
```

Now, we will save the reformatted `df` into a new rds file called `part0.rds`. The output will be saved into a sub-folder called rds.

```{r}
#| eval: false
write_rds(df,"../../data/rds/In-class_Ex/part0.rds")
```

rds is a R data file format. Saving it as so will keep the data's structure.

## Extracting trip starting locations

-   extracting trips' origin locations

-   derive three new columns (i.e. variables) for weekday, starting hour and day of the month

-   name the output tibble data.frame `origin_df`

```{r}
#| eval: false
origin_df <- df %>%
  group_by(trj_id) %>%
  arrange(pingtimestamp) %>% # from dplyr package
  filter(row_number()==1) %>%
  mutate(weekday = wday(pingtimestamp, # functions from lubridate package
                        label=TRUE,
                        abbr=TRUE),
         start_hr = factor(hour(pingtimestamp)), # With factor(), it makes it ordinal
         day = factor(mday(pingtimestamp)))
view(origin_df)
```

## Extract trip ending locations

-   extracting trips' ending locations

-   derive three new columns for weekday, ending hour and day of the month

-   name the output tibble data.frame `destination_df`

```{r}
#| eval: false
destination_df <- df %>%
  group_by(trj_id) %>%
  arrange(desc(pingtimestamp)) %>% # from dplyr package
  filter(row_number()==1) %>%
  mutate(weekday = wday(pingtimestamp, # functions from lubridate package
                        label=TRUE,
                        abbr=TRUE),
         end_hr = factor(hour(pingtimestamp)), # With factor(), it makes it ordinal
         day = factor(mday(pingtimestamp)))
view(destination_df)
```

## Save the extracted data

```{r}
#| eval: false
write_rds(origin_df,"../../data/rds/In-class_Ex/origin_df.rds")
write_rds(destination_df,"../../data/rds/In-class_Ex/destination_df.rds")
```

## Import saved data

```{r}
origin_df <- read_rds("../../data/rds/In-class_Ex/origin_df.rds")
destination_df <- read_rds("../../data/rds/In-class_Ex/destination_df.rds")
```

### Visualizing frequency distribution

# Converting into sf tibble data.frame

# Visualizing as Point Symbol Map

# Importing Master Plan 2019 Planning Subzone Boundary
