---
title: "Prototyping Modules for Geospatial Analytics Shiny Application"
subtitle: "Take-home_Ex03"
format:
  html:
    code-fold: false
    code-summary: "Show the code"
    toc: true # Table of Contents
execute:
  eval: true
  echo: true
  freeze: true
  warning: False  # This turns off warning messages being displayed
date: 20 March, 2024
date-modified: "last-modified"
---

# 1 Introduction

## 1.1 The Objective

The purpose of this is to choose a module of the proposed Geospatial Analytics Shiny Application and complete the following tasks:

-   To evaluate and determine the necessary R packages needed for the Shiny application are supported in R CRAN

-   To prepare and test the specific R codes can be run and return the correct output as expected

-   To determine the parameters and outputs that will be exposed on the Shiny applications

-   To select the appropriate Shiny UI components for exposing the parameters determined above

## 1.2 The Selected Module

The module that I have chosen to work on is "Modeling Geographic Accessibility". The following chapters will show the code used for this module and its output.

The purpose of this module is to map out the **geographic accessibility of different regions in Singapore, taking into account various amenities**, such as schools, MRT stations, malls, supermarkets and hawkers.

# 2 The Data

The data we have chosen has been compiled from multiple sources. For each amenity type, we have alist of names that have been web-scrapped or manually collected from the internet (e.g. school names from MOE). The list of names will then be put through a python API ([onemapsg](https://pypi.org/project/onemapsg/)) which will take a location name as input and output its coordinates in WGS84 format.

The dataset consists of various location names and coordinates under these amenities types:

-   Schools (Pri, Sec, JC, Poly, ITE, Kindergarten)
-   MRT
-   Malls
-   Supermarket
-   Hawkers

We will also be using:

-   `MP14_SUBZONE_NO_SEA_PL`: URA Master Plan 2014 subzone boundary GIS data. This data set is downloaded from data.gov.sg.

-   `hexagons`: A 250m radius hexagons GIS data. This data set was created by using [*st_make_grid()*](https://r-spatial.github.io/sf/reference/st_make_grid.html) of sf package. It is in ESRI shapefile format.

# 3 Getting Started

```{r}
pacman::p_load(tmap, SpatialAcc, sf, ggstatsplot, reshape2, tidyverse)
```

# 4 Data Wrangling

## 4.1 Importing geospatial data

We will import the: `mpsz` and `hexagons` data and make sure that it is in the correct format.

```{r}
mpsz <- st_read(dsn="data/geospatial", layer = "MP14_SUBZONE_NO_SEA_PL") %>%
  st_transform(3414)
```

```{r}
hexagons <- st_read(dsn = "data/geospatial", layer="hexagons") %>%
  st_transform(3414)
```

## 4.2 Aspatial Data Wrangling

We will import the amenities data based on type.

### 4.2.1 Schools Data

```{r}
schools <- read_csv("data/aspatial/school_coordinates.csv")
schools_sf <- st_as_sf(schools,
                       coords = c("Longitude", "Latitude"),
                       crs=4326) %>%
  st_transform(3414)
```

```{r}
schools_sf
```

### 4.2.2 MRT Data

```{r}
MRT <- read_csv("data/aspatial/MRT_coordinates.csv")
MRT_sf <- st_as_sf(MRT,
                       coords = c("Longitude", "Latitude"),
                       crs=4326) %>%
  st_transform(3414)
```

```{r}
MRT_sf
```

### 4.2.3 Mall Data

```{r}
mall <- read_csv("data/aspatial/mall_coordinates.csv")
mall_sf <- st_as_sf(mall,
                       coords = c("Longitude", "Latitude"),
                       crs=4326) %>%
  st_transform(3414)
```

```{r}
mall_sf
```

### 4.2.4 Supermarket Data

```{r}
supermarket <- read_csv("data/aspatial/supermarket_coordinates.csv")
supermarket_sf <- st_as_sf(supermarket,
                       coords = c("Longitude", "Latitude"),
                       crs=4326) %>%
  st_transform(3414)
```

```{r}
supermarket_sf
```

### 4.2.5 Hawker Data

```{r}
hawker <- read_csv("data/aspatial/hawker_coordinates.csv")
hawker_sf <- st_as_sf(hawker,
                       coords = c("Longitude", "Latitude"),
                       crs=4326) %>%
  st_transform(3414)
```

```{r}
hawker_sf
```

## 4.3 Cleaning and updating attribute fields of the geospatial data

For each amenity, we will specify the capacity, for each hexagon, we will specify the demand.

```{r}
schools_sf <- schools_sf %>% 
  select(...1, SCHOOL, TYPE) %>%
  mutate(capacity = 100)

MRT_sf <- MRT_sf %>%
  mutate(capacity = 100)

mall_sf <- mall_sf %>%
  mutate(capacity = 100)

supermarket_sf <- supermarket_sf %>%
  mutate(capacity = 100)

hawker_sf <- hawker_sf %>%
  mutate(capacity = 100)
```

```{r}
hexagons <- hexagons %>%
  select(fid) %>%
  mutate(demand = 100)
```

## 4.4 Getting centroid points from hexagon

We need to get the centroid of the hexagon so that we can use it to calculate the euclidean distance from each amenity location later.

```{r}
hexagon_points <- st_centroid(hexagons)
```

```{r}
hexagon_points
```

## 4.5 Get Distance Matrix

Based on the hexagon centroids and location of the schools, we will generate a distance matrix.

### 4.5.1 Schools Distance Matrix

```{r}
schools_distance_matrix <- st_distance(hexagon_points, schools_sf, by_element = FALSE) 
schools_distance_matrix_km <- schools_distance_matrix / 1000 
schools_distance_matrix_km <- matrix(as.numeric(schools_distance_matrix_km), nrow = nrow(schools_distance_matrix_km), ncol = ncol(schools_distance_matrix_km))
```

### 4.5.2 MRT Distance Matrix

```{r}
MRT_distance_matrix <- st_distance(hexagon_points, MRT_sf, by_element = FALSE) 
MRT_distance_matrix_km <- MRT_distance_matrix / 1000 
MRT_distance_matrix_km <- matrix(as.numeric(MRT_distance_matrix_km), nrow = nrow(MRT_distance_matrix_km), ncol = ncol(MRT_distance_matrix_km))
```

### 4.5.3 Mall Distance Matrix

```{r}
mall_distance_matrix <- st_distance(hexagon_points, mall_sf, by_element = FALSE) 
mall_distance_matrix_km <- mall_distance_matrix / 1000 
mall_distance_matrix_km <- matrix(as.numeric(mall_distance_matrix_km), nrow = nrow(mall_distance_matrix_km), ncol = ncol(mall_distance_matrix_km))
```

### 4.5.4 Supermarket Distance Matrix

```{r}
supermarket_distance_matrix <- st_distance(hexagon_points, supermarket_sf, by_element = FALSE) 
supermarket_distance_matrix_km <- supermarket_distance_matrix / 1000 
supermarket_distance_matrix_km <- matrix(as.numeric(supermarket_distance_matrix_km), nrow = nrow(supermarket_distance_matrix_km), ncol = ncol(supermarket_distance_matrix_km))
```

### 4.5.5 Hawker Distance Matrix

```{r}
hawker_distance_matrix <- st_distance(hexagon_points, hawker_sf, by_element = FALSE) 
hawker_distance_matrix_km <- hawker_distance_matrix / 1000 
hawker_distance_matrix_km <- matrix(as.numeric(hawker_distance_matrix_km), nrow = nrow(hawker_distance_matrix_km), ncol = ncol(hawker_distance_matrix_km))
```

# 5 Exploratory Data Analysis (EDA)

Let's start by counting the number of schools in each region.

```{r}
mpsz$`Sch Count`<- lengths(st_intersects(mpsz, schools_sf))
```

To list the zone with the most number of schools:

```{r}
top_n(mpsz, 1, `Sch Count`)
```

We will also calculate the density of schools in that area:

```{r}
mpsz$Area <- mpsz %>%
  st_area()

mpsz <- mpsz %>%
  mutate(`Sch Density` = `Sch Count`/Area *1000000)
```

```{r}
ggplot(data=mpsz, 
       aes(x= as.numeric(`Sch Density`)))+
  geom_histogram(bins=10, 
                 color="black", 
                 fill="light blue") +
  labs(title = "Are schools evenly distributed in Singapore?",
       subtitle= "Most sub-zones only have one school.",
      x = "School density (per km sq)",
      y = "Frequency")
```

# 7 Analyze Accessibility using Hansen Method

## 7.1 Computing Hansen's Accessibility

### 7.1.1 School Hansen's Accessibility

```{r}
schools_acc_Hansen <- data.frame(ac(hexagons$demand,
                            schools_sf$capacity,
                            D = schools_distance_matrix_km, 
                            d0 = 50,
                            power = 2, 
                            family = "Hansen"))
colnames(schools_acc_Hansen) <- "schools_accHansen"
schools_acc_Hansen <- as_tibble(schools_acc_Hansen)
```

### 7.1.2 MRT Hansen's Accessibility

```{r}
MRT_acc_Hansen <- data.frame(ac(hexagons$demand,
                            MRT_sf$capacity,
                            D = MRT_distance_matrix_km, 
                            d0 = 50,
                            power = 2, 
                            family = "Hansen"))
colnames(MRT_acc_Hansen) <- "MRT_accHansen"
MRT_acc_Hansen <- as_tibble(MRT_acc_Hansen)
```

### 7.1.3 Mall Hansen's Accessibility

```{r}
mall_acc_Hansen <- data.frame(ac(hexagons$demand,
                            mall_sf$capacity,
                            D = mall_distance_matrix_km, 
                            d0 = 50,
                            power = 2, 
                            family = "Hansen"))
colnames(mall_acc_Hansen) <- "mall_accHansen"
mall_acc_Hansen <- as_tibble(mall_acc_Hansen)
```

### 7.1.4 Supermarket Hansen's Accessibility

```{r}
supermarket_acc_Hansen <- data.frame(ac(hexagons$demand,
                            supermarket_sf$capacity,
                            D = supermarket_distance_matrix_km, 
                            d0 = 50,
                            power = 2, 
                            family = "Hansen"))
colnames(supermarket_acc_Hansen) <- "supermarket_accHansen"
supermarket_acc_Hansen <- as_tibble(supermarket_acc_Hansen)
```

### 7.1.5 Hawker Hansen's Accessibility

```{r}
hawker_acc_Hansen <- data.frame(ac(hexagons$demand,
                            hawker_sf$capacity,
                            D = hawker_distance_matrix_km, 
                            d0 = 50,
                            power = 2, 
                            family = "Hansen"))
colnames(hawker_acc_Hansen) <- "hawker_accHansen"
hawker_acc_Hansen <- as_tibble(hawker_acc_Hansen)
```

### 7.1.6 Combine all Hansen's Accessibility into one data frame

Now, we will bind all the acc_Hansen tibble data frame with the hexagons simple feature data frame.

```{r}
hexagon_Hansen <- bind_cols(hexagons, schools_acc_Hansen, MRT_acc_Hansen, mall_acc_Hansen, supermarket_acc_Hansen, hawker_acc_Hansen)
```

## 7.2 Visualizing Hansen's Accessibility

```{r}
mapex <- st_bbox(hexagons)
```

Plotting Hansen's Accessibility for all amenities

### 7.2.1 Schools

```{r}
tmap_mode("plot")
tm_shape(hexagon_Hansen,
         bbox = mapex) + 
  tm_fill(col = "schools_accHansen",
          n = 10,
          style = "quantile",
          border.col = "black",
          border.lwd = 1) +
tm_shape(schools_sf) +
  tm_symbols(size = 0.1) +
  tm_layout(main.title = "Accessibility to Schools: Hansen method",
            main.title.position = "center",
            main.title.size = 2,
            legend.outside = FALSE,
            legend.height = 0.45, 
            legend.width = 3.0,
            legend.format = list(digits = 6),
            legend.position = c("right", "top"),
            frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar(width = 0.15) +
  tm_grid(lwd = 0.1, alpha = 0.5)
```

### 7.2.2 MRTs

```{r}
tmap_mode("plot")
tm_shape(hexagon_Hansen,
         bbox = mapex) + 
  tm_fill(col = "MRT_accHansen",
          n = 10,
          style = "quantile",
          border.col = "black",
          border.lwd = 1) +
tm_shape(MRT_sf) +
  tm_symbols(size = 0.1) +
  tm_layout(main.title = "Accessibility to MRTs: Hansen method",
            main.title.position = "center",
            main.title.size = 2,
            legend.outside = FALSE,
            legend.height = 0.45, 
            legend.width = 3.0,
            legend.format = list(digits = 6),
            legend.position = c("right", "top"),
            frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar(width = 0.15) +
  tm_grid(lwd = 0.1, alpha = 0.5)
```

### 7.2.3 Malls

```{r}
tmap_mode("plot")
tm_shape(hexagon_Hansen,
         bbox = mapex) + 
  tm_fill(col = "mall_accHansen",
          n = 10,
          style = "quantile",
          border.col = "black",
          border.lwd = 1) +
tm_shape(mall_sf) +
  tm_symbols(size = 0.1) +
  tm_layout(main.title = "Accessibility to Malls: Hansen method",
            main.title.position = "center",
            main.title.size = 2,
            legend.outside = FALSE,
            legend.height = 0.45, 
            legend.width = 3.0,
            legend.format = list(digits = 6),
            legend.position = c("right", "top"),
            frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar(width = 0.15) +
  tm_grid(lwd = 0.1, alpha = 0.5)
```

### 7.2.4 Supermarkets

```{r}
tmap_mode("plot")
tm_shape(hexagon_Hansen,
         bbox = mapex) + 
  tm_fill(col = "supermarket_accHansen",
          n = 10,
          style = "quantile",
          border.col = "black",
          border.lwd = 1) +
tm_shape(supermarket_sf) +
  tm_symbols(size = 0.1) +
  tm_layout(main.title = "Accessibility to Supermarkets: Hansen method",
            main.title.position = "center",
            main.title.size = 2,
            legend.outside = FALSE,
            legend.height = 0.45, 
            legend.width = 3.0,
            legend.format = list(digits = 6),
            legend.position = c("right", "top"),
            frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar(width = 0.15) +
  tm_grid(lwd = 0.1, alpha = 0.5)
```

### 7.2.5 Hawkers

```{r}
tmap_mode("plot")
tm_shape(hexagon_Hansen,
         bbox = mapex) + 
  tm_fill(col = "hawker_accHansen",
          n = 10,
          style = "quantile",
          border.col = "black",
          border.lwd = 1) +
tm_shape(hawker_sf) +
  tm_symbols(size = 0.1) +
  tm_layout(main.title = "Accessibility to Hawkers: Hansen method",
            main.title.position = "center",
            main.title.size = 2,
            legend.outside = FALSE,
            legend.height = 0.45, 
            legend.width = 3.0,
            legend.format = list(digits = 6),
            legend.position = c("right", "top"),
            frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar(width = 0.15) +
  tm_grid(lwd = 0.1, alpha = 0.5)
```

## 7.3 Statistical Graphic Visualization

Now, let's show an example of how the code will look when using a more statistical visualization.

```{r}
hexagon_Hansen <- st_join(hexagon_Hansen, mpsz, join = st_intersects)
```

```{r}
ggplot(data=hexagon_Hansen, 
       aes(y = log(schools_accHansen), 
           x = REGION_N)) +
  geom_boxplot() +
  geom_point(stat="summary", 
             fun.y="mean", 
             colour ="red", 
             size=2)
```

# 8 Analyze Accessibility using KD2SFCA Method

## 8.1 Computing KD2SFCA's Accessibility

### 8.1.1 School KD2SFCA's Accessibility

```{r}
schools_acc_KD2SFCA <- data.frame(ac(hexagons$demand,
                            schools_sf$capacity,
                            schools_distance_matrix_km,
                            d0 = 50,
                            power = 2, 
                            family = "KD2SFCA"))

colnames(schools_acc_KD2SFCA) <- "schools_accKD2SFCA"
schools_acc_KD2SFCA <- as_tibble(schools_acc_KD2SFCA)
```

### 8.1.2 MRT KD2SFCA's Accessibility

```{r}
MRT_acc_KD2SFCA <- data.frame(ac(hexagons$demand,
                            MRT_sf$capacity,
                            MRT_distance_matrix_km,
                            d0 = 50,
                            power = 2, 
                            family = "KD2SFCA"))

colnames(MRT_acc_KD2SFCA) <- "MRT_accKD2SFCA"
MRT_acc_KD2SFCA <- as_tibble(MRT_acc_KD2SFCA)
```

### 8.1.3 Mall KD2SFCA's Accessibility

```{r}
mall_acc_KD2SFCA <- data.frame(ac(hexagons$demand,
                            mall_sf$capacity,
                            mall_distance_matrix_km,
                            d0 = 50,
                            power = 2, 
                            family = "KD2SFCA"))

colnames(mall_acc_KD2SFCA) <- "mall_accKD2SFCA"
mall_acc_KD2SFCA <- as_tibble(mall_acc_KD2SFCA)
```

### 8.1.4 Supermarket KD2SFCA's Accessibility

```{r}
supermarket_acc_KD2SFCA <- data.frame(ac(hexagons$demand,
                            supermarket_sf$capacity,
                            supermarket_distance_matrix_km,
                            d0 = 50,
                            power = 2, 
                            family = "KD2SFCA"))

colnames(supermarket_acc_KD2SFCA) <- "supermarket_accKD2SFCA"
supermarket_acc_KD2SFCA <- as_tibble(supermarket_acc_KD2SFCA)
```

### 8.1.5 Hawker KD2SFCA's Accessibility

```{r}
hawker_acc_KD2SFCA <- data.frame(ac(hexagons$demand,
                            hawker_sf$capacity,
                            hawker_distance_matrix_km,
                            d0 = 50,
                            power = 2, 
                            family = "KD2SFCA"))

colnames(hawker_acc_KD2SFCA) <- "hawker_accKD2SFCA"
hawker_acc_KD2SFCA <- as_tibble(hawker_acc_KD2SFCA)
```

### 8.1.1 Combine all KD2SFCA's Accessibility into one data frame

```{r}
hexagon_KD2SFCA <- bind_cols(hexagons, schools_acc_KD2SFCA, MRT_acc_KD2SFCA, mall_acc_KD2SFCA, supermarket_acc_KD2SFCA, hawker_acc_KD2SFCA)
```

## 8.2 Visualizing KD2SFCA's Accessibility

Plotting KD2SFCA's Accessibility for all amenities

### 8.2.1 Schools

```{r}
tmap_mode("plot")
tm_shape(hexagon_KD2SFCA,
         bbox = mapex) + 
  tm_fill(col = "schools_accKD2SFCA",
          n = 10,
          style = "quantile",
          palette="Greens",
          border.col = "black",
          border.lwd = 1) +
tm_shape(schools_sf) +
  tm_symbols(size = 0.1) +
  tm_layout(main.title = "Accessibility to Schools: KD2SFCA method",
            main.title.position = "center",
            main.title.size = 2,
            legend.outside = FALSE,
            legend.height = 0.45, 
            legend.width = 3.0,
            legend.format = list(digits = 6),
            legend.position = c("right", "top"),
            frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar(width = 0.15) +
  tm_grid(lwd = 0.1, alpha = 0.5)
```

### 8.2.2 MRTs

```{r}
tmap_mode("plot")
tm_shape(hexagon_KD2SFCA,
         bbox = mapex) + 
  tm_fill(col = "MRT_accKD2SFCA",
          n = 10,
          style = "quantile",
          palette="Greens",
          border.col = "black",
          border.lwd = 1) +
tm_shape(MRT_sf) +
  tm_symbols(size = 0.1) +
  tm_layout(main.title = "Accessibility to MRT: KD2SFCA method",
            main.title.position = "center",
            main.title.size = 2,
            legend.outside = FALSE,
            legend.height = 0.45, 
            legend.width = 3.0,
            legend.format = list(digits = 6),
            legend.position = c("right", "top"),
            frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar(width = 0.15) +
  tm_grid(lwd = 0.1, alpha = 0.5)
```

### 8.2.3 Malls

```{r}
tmap_mode("plot")
tm_shape(hexagon_KD2SFCA,
         bbox = mapex) + 
  tm_fill(col = "mall_accKD2SFCA",
          n = 10,
          style = "quantile",
          palette="Greens",
          border.col = "black",
          border.lwd = 1) +
tm_shape(mall_sf) +
  tm_symbols(size = 0.1) +
  tm_layout(main.title = "Accessibility to Malls: KD2SFCA method",
            main.title.position = "center",
            main.title.size = 2,
            legend.outside = FALSE,
            legend.height = 0.45, 
            legend.width = 3.0,
            legend.format = list(digits = 6),
            legend.position = c("right", "top"),
            frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar(width = 0.15) +
  tm_grid(lwd = 0.1, alpha = 0.5)
```

### 8.2.4 Supermarkets

```{r}
tmap_mode("plot")
tm_shape(hexagon_KD2SFCA,
         bbox = mapex) + 
  tm_fill(col = "supermarket_accKD2SFCA",
          n = 10,
          style = "quantile",
          palette="Greens",
          border.col = "black",
          border.lwd = 1) +
tm_shape(supermarket_sf) +
  tm_symbols(size = 0.1) +
  tm_layout(main.title = "Accessibility to Supermarkets: KD2SFCA method",
            main.title.position = "center",
            main.title.size = 2,
            legend.outside = FALSE,
            legend.height = 0.45, 
            legend.width = 3.0,
            legend.format = list(digits = 6),
            legend.position = c("right", "top"),
            frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar(width = 0.15) +
  tm_grid(lwd = 0.1, alpha = 0.5)
```

### 8.2.5 Hawkers

```{r}
tmap_mode("plot")
tm_shape(hexagon_KD2SFCA,
         bbox = mapex) + 
  tm_fill(col = "hawker_accKD2SFCA",
          n = 10,
          style = "quantile",
          palette="Greens",
          border.col = "black",
          border.lwd = 1) +
tm_shape(hawker_sf) +
  tm_symbols(size = 0.1) +
  tm_layout(main.title = "Accessibility to Hawkers: KD2SFCA method",
            main.title.position = "center",
            main.title.size = 2,
            legend.outside = FALSE,
            legend.height = 0.45, 
            legend.width = 3.0,
            legend.format = list(digits = 6),
            legend.position = c("right", "top"),
            frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar(width = 0.15) +
  tm_grid(lwd = 0.1, alpha = 0.5)
```

# 9 Analyze Accessibility using Spatial Accessibility Measure (SAM) Method

## 9.1 Computing SAM Accessibility

### 9.1.1 School SAM Accessibility

```{r}
schools_acc_SAM <- data.frame(ac(hexagons$demand,
                            schools_sf$capacity,
                            D = schools_distance_matrix_km, 
                            d0 = 50,
                            power = 2, 
                            family = "SAM"))
colnames(schools_acc_SAM) <- "schools_accSAM"
schools_acc_SAM <- as_tibble(schools_acc_SAM)
```

### 9.1.2 MRT SAM Accessibility

```{r}
MRT_acc_SAM <- data.frame(ac(hexagons$demand,
                            MRT_sf$capacity,
                            D = MRT_distance_matrix_km, 
                            d0 = 50,
                            power = 2, 
                            family = "SAM"))
colnames(MRT_acc_SAM) <- "MRT_accSAM"
MRT_acc_SAM <- as_tibble(MRT_acc_SAM)
```

### 9.1.3 Mall SAM Accessibility

```{r}
mall_acc_SAM <- data.frame(ac(hexagons$demand,
                            mall_sf$capacity,
                            D = mall_distance_matrix_km, 
                            d0 = 50,
                            power = 2, 
                            family = "SAM"))
colnames(mall_acc_SAM) <- "mall_accSAM"
mall_acc_SAM <- as_tibble(mall_acc_SAM)
```

### 9.1.4 Supermarket SAM Accessibility

```{r}
supermarket_acc_SAM <- data.frame(ac(hexagons$demand,
                            supermarket_sf$capacity,
                            D = supermarket_distance_matrix_km, 
                            d0 = 50,
                            power = 2, 
                            family = "SAM"))
colnames(supermarket_acc_SAM) <- "supermarket_accSAM"
supermarket_acc_SAM <- as_tibble(supermarket_acc_SAM)
```

### 9.1.5 Hawker SAM Accessibility

```{r}
hawker_acc_SAM <- data.frame(ac(hexagons$demand,
                            hawker_sf$capacity,
                            D = hawker_distance_matrix_km, 
                            d0 = 50,
                            power = 2, 
                            family = "SAM"))
colnames(hawker_acc_SAM) <- "hawker_accSAM"
hawker_acc_SAM <- as_tibble(hawker_acc_SAM)
```

### 9.1.6 Combine all SAM Accessibility into one data frame

```{r}
hexagon_SAM <- bind_cols(hexagons, schools_acc_SAM, MRT_acc_SAM, mall_acc_SAM, supermarket_acc_SAM, hawker_acc_SAM)
```

## 9.2 Visualizing SAM's Accessibility

### 9.2.1 Schools

```{r}
tmap_mode("plot")
tm_shape(hexagon_SAM,
         bbox = mapex) + 
  tm_fill(col = "schools_accSAM",
          n = 10,
          style = "quantile",
          palette = "Blues",
          border.col = "black",
          border.lwd = 1) +
tm_shape(schools_sf) +
  tm_symbols(size = 0.1) +
  tm_layout(main.title = "Accessibility to Schools: SAM method",
            main.title.position = "center",
            main.title.size = 2,
            legend.outside = FALSE,
            legend.height = 0.45, 
            legend.width = 3.0,
            legend.format = list(digits = 3),
            legend.position = c("right", "top"),
            frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar(width = 0.15) +
  tm_grid(lwd = 0.1, alpha = 0.5)
```

### 9.2.2 MRTs

```{r}
tmap_mode("plot")
tm_shape(hexagon_SAM,
         bbox = mapex) + 
  tm_fill(col = "MRT_accSAM",
          n = 10,
          style = "quantile",
          palette = "Blues",
          border.col = "black",
          border.lwd = 1) +
tm_shape(MRT_sf) +
  tm_symbols(size = 0.1) +
  tm_layout(main.title = "Accessibility to MRTs: SAM method",
            main.title.position = "center",
            main.title.size = 2,
            legend.outside = FALSE,
            legend.height = 0.45, 
            legend.width = 3.0,
            legend.format = list(digits = 3),
            legend.position = c("right", "top"),
            frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar(width = 0.15) +
  tm_grid(lwd = 0.1, alpha = 0.5)
```

### 9.2.3 Malls

```{r}
tmap_mode("plot")
tm_shape(hexagon_SAM,
         bbox = mapex) + 
  tm_fill(col = "mall_accSAM",
          n = 10,
          style = "quantile",
          palette = "Blues",
          border.col = "black",
          border.lwd = 1) +
tm_shape(mall_sf) +
  tm_symbols(size = 0.1) +
  tm_layout(main.title = "Accessibility to Malls: SAM method",
            main.title.position = "center",
            main.title.size = 2,
            legend.outside = FALSE,
            legend.height = 0.45, 
            legend.width = 3.0,
            legend.format = list(digits = 3),
            legend.position = c("right", "top"),
            frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar(width = 0.15) +
  tm_grid(lwd = 0.1, alpha = 0.5)
```

### 9.2.4 Supermarkets

```{r}
tmap_mode("plot")
tm_shape(hexagon_SAM,
         bbox = mapex) + 
  tm_fill(col = "supermarket_accSAM",
          n = 10,
          style = "quantile",
          palette = "Blues",
          border.col = "black",
          border.lwd = 1) +
tm_shape(supermarket_sf) +
  tm_symbols(size = 0.1) +
  tm_layout(main.title = "Accessibility to Supermarkets: SAM method",
            main.title.position = "center",
            main.title.size = 2,
            legend.outside = FALSE,
            legend.height = 0.45, 
            legend.width = 3.0,
            legend.format = list(digits = 3),
            legend.position = c("right", "top"),
            frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar(width = 0.15) +
  tm_grid(lwd = 0.1, alpha = 0.5)
```

### 9.2.5 Hawkers

```{r}
tmap_mode("plot")
tm_shape(hexagon_SAM,
         bbox = mapex) + 
  tm_fill(col = "hawker_accSAM",
          n = 10,
          style = "quantile",
          palette = "Blues",
          border.col = "black",
          border.lwd = 1) +
tm_shape(hawker_sf) +
  tm_symbols(size = 0.1) +
  tm_layout(main.title = "Accessibility to Hawkers: SAM method",
            main.title.position = "center",
            main.title.size = 2,
            legend.outside = FALSE,
            legend.height = 0.45, 
            legend.width = 3.0,
            legend.format = list(digits = 3),
            legend.position = c("right", "top"),
            frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar(width = 0.15) +
  tm_grid(lwd = 0.1, alpha = 0.5)
```

# 10 Shiny Application UI Design

The proposed output of this module will be a dashboard of the all the accessibility heat maps. It will feature:

-   A compilation of all 5 maps side-by-side.

-   A filter for accessibility methods. To display by either Hansen, KD2SFCA or SAM method.

-   A toggle to display only the selected Amenity type and all it's accessibility based on all methods. The statistical visualization (box plot) can also be shown alongside here.

-   Possibly use another package with more interactive functions such as to show the details of each amenity point when hovering over it (such as the name of the school and its type)

# 11 Interactive Maps for Dashboards

Here, we experiment with plotting the School SAM accessibility scores on an interactive map that shows the amenity details when clicked on. Try clicking on the red points on the map, it should show the School name and Type.

```{r}
pacman::p_load(leaflet, scales, sf)
```

```{r}
# Transform data into WGS84 since leaflet expects it
schools_sf_wgs84 <- st_transform(schools_sf, 4326)
hexagon_SAM_wgs84 <- st_transform(hexagon_SAM, 4326)


# Function to assign colors based on quantiles
assignColors <- function(x, n = 5, palette = "Blues") {
  # Calculate quantiles
  bins <- quantile(x, probs = seq(0, 1, length.out = n + 1), na.rm = TRUE)
  # Create color palette
  colors <- colorBin(palette = palette, domain = x, bins = bins, na.color = NA)(x)
  return(colors)
}

# Apply the function to your accessibility scores
hex_colors <- assignColors(hexagon_SAM_wgs84$schools_accSAM, n = 10, palette = "Blues")

# Create the leaflet map with manual color assignments
m <- leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  
  # Add hexagons with manually assigned colors
  addPolygons(data = hexagon_SAM_wgs84, fillColor = hex_colors,
              fillOpacity = 1, color = NA, weight = 0, smoothFactor = 0) %>%
  
  # Adjust schools' visualization as before
  addCircleMarkers(data = schools_sf_wgs84, radius = 1, color = "red", fillColor = "red",
                   fillOpacity = 1, popup = ~paste("<b>SCHOOL:</b>", SCHOOL, "<br><b>TYPE:</b>", TYPE)) %>%
  
  setView(lng = mean(st_coordinates(schools_sf_wgs84)[,1]), lat = mean(st_coordinates(schools_sf_wgs84)[,2]), zoom = 12)

# Print the map
m
```
