---
title: "Global and Local Measures of Spatial Autocorrelation - sfdep methods"
subtitle: "In-class Exercise 5"
format:
  html:
    code-fold: false
    code-summary: "Show the code"
    toc: true # Table of Contents
execute:
  eval: true
  echo: true
  warning: False  # This turns off warning messages being displayed
date: 5 February, 2024
date-modified: "last-modified"
---

# Getting Started

## Installing and Loading the R packages

Four R packages will be used for this in-class exercise, they are: sf, sfdep, tmap and tidyverse

```{r}
pacman::p_load(sf, sfdep, tmap, tidyverse)
```

# The Data

For the purpose of this in-class exercise, the Hunan data sets will be used. There are two data sets in this use case, they are:

-   Hunan, a geospatial data set in ESRI shapefile format, and

-   Hunan_2012, an attribute data set in csv format

## Importing geospatial data

```{r}
hunan <- st_read(dsn="../../data/geospatial/HunanBoundary",
                 layer="Hunan")
```

## Importing attribute table

```{r}
hunan2012 <- read_csv("../../data/aspatial/Hunan/Hunan_2012.csv")
```

## Combining both data frame by using left join

```{r}
hunan_GDPPC <- left_join(hunan,hunan2012) %>%
  select(1:4, 7, 15)
```

## Plotting a Choropleth Map

```{r}
tmap_mode("plot")
tm_shape(hunan_GDPPC) +
  tm_fill("GDPPC",
          style = 'quantile',
          palette = 'Blues',
          title = 'GDPPC') +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title="Distribution of GDP per capita by county, Hunan Province",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45,
            legend.width = 0.35,
            frame = TRUE) +
  tm_compass(type='8star', size=2) +
  tm_scale_bar() +
  tm_grid(alpha = 0.2)
```

```{r}
wm_q <- hunan_GDPPC %>%
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(nb,
                         style = "W"),
         .before = 1)
```

```{r}
moranI <- global_moran(wm_q$GDPPC,
                       wm_q$nb,
                       wm_q$wt)
glimpse(moranI)
```
